Temporal ordering and registration of images in studies of developmental dynamics

### 2014-09-01 image_analysis_paper2/pnas_draft.tex
added stas comments

\significancetext{
Spatial registration and temporal ordering of images of developing embryos are one of the first steps in the reconstruction of developmental dynamics. 
%
When the number of images is small and the differences between them are easily recognizable, image registration and ordering can be done manually; these tasks become challenging when the data sets are larger and more complex. 
%
Here we show this can be automated and combined in a single
step using recent advances in data mining. 
%
To illustrate this approach, we temporally order and register data sets collected
in studies of cell signaling in the early fruit fly embryo.}

@@

Imaging studies provide unique insights into the dynamics of pattern formation and morphogenesis in developing tissues.
%
In one of the most commonly used experimental formats, developmental progress is reconstructed from snapshots of chemical or mechanical processes in fixed embryos.
%
As a first step in these reconstructions, such snapshots must be spatially registered and ordered in time.
%
Currently, registration and ordering of images is often done manually, requiring a significant amount of expertise with a specific developmental system.
%
However, as the sizes of imaging data sets grow, these tasks become increasingly difficult, especially when the images are noisy and the examined developmental changes are subtle.
%
To address these challenges, we present an automated approach to simultaneously register and temporally order imaging data sets.
%
The approach is based on vector diffusion maps, a broadly applicable manifold learning technique that does not require {\it a priori} knowledge of image features or a parametric model of the developmental dynamics.
%
We validate our approach using a live imaging data set where the correct rotational orientation and temporal order are known.
%
Finally, we demonstrate that our approach can successfully extract a representative developmental trajectory from a collection of images of fixed {\it Drosophila} embryos.

@@

We present a robust algorithmic approach to registration and temporal ordering.
%
In contrast to a number of previous methodologies \cite{zitova2003image, rowley1998rotation, hajnal2010medical, greenspan1994rotation, zhao2003face}, our methodology does not rely on the {\em a priori} knowledge of spatial landmarks for registration or markers of developmental progression for temporal ordering.
%
Our approach is based on a manifold learning algorithm (vector diffusion maps \cite{singer2012vector}) which simultaneously addresses the problems of registration and temporal ordering. 
%
This algorithm is one of several nonlinear dimensionality reduction techniques that have been developed over the past decade \cite{Belkin2003, coifman2005geometric, coifman2006geometric, tenenbaum2000global, roweis2000nonlinear}, for
applications ranging from analysis of cryo-electron microscopy (cryo-EM) images of individual molecules  \cite{zhao2014rotationally, singer2011viewing} to face recognition \cite{lafon2006data}.
%
Here, the vector diffusion maps algorithm is adapted for the analysis of images of developing embryos in studies of developmental dynamics, with the main objective of revealing stereotypic developmental trajectories from fixed images.
%
To illustrate our approach, we analyze two data sets from a study of {\it Drosophila} embryogenesis, one of the best experimental models for studies of development \cite{jaeger2012drosophila}.
%
Our first data set, from a live imaging study where the correct rotational orientation and temporal order are independently known, will be used to validate our approach.
%
Our second data set consists of images from fixed embryos where the correct orientation and order is unknown; here, we will show how the algorithm can help uncover developmental dynamics which are not readily apparent.

@@

After removing variability due to rotations, the developmental dynamics may be revealed by ordering the data along the one-dimensional manifold that parameterizes most of the remaining variability in the data.
%
Such a manifold can be discovered using diffusion maps \cite{coifman2005geometric}, a nonlinear dimensionality reduction technique that uncovers an intrinsic parametrization of data that lies on a low-dimensional, perhaps nonlinear, manifold in high-dimensional space.
%
The idea is illustrated in \fig~\ref{subfig:dmaps1}, where the data are two-dimensional points which lie on a one-dimensional nonlinear curve.
%
We use {\it local} information about the data to find a parametrization  which respects the underlying manifold geometry, so that points which are close in high-dimensional space (e.g., images which look similar) are close in our parametrization.
%
This idea of locality is denoted by the color of the edges in \fig~\ref{subfig:dmaps1}:
data points which are close are connected by dark edges, and clearly, the dark edges are more ``informative" about the low-dimensional structure of the data.
%
The color in \fig~\ref{subfig:dmaps2} depicts the one-dimensional parametrization or ordering of the data that we can detect visually.
%
In our working examples, each data point will be of much higher dimension (e.g., a pixelated image), and so we cannot extract this low-dimensional structure visually.
%
Instead, we will use diffusion maps to automatically uncover a parametrization of our high-dimensional data (see \SI).
%
Diffusion maps generalizes directly from one-dimensional nonlinear curves to higher-dimensional manifolds.

@@

To validate the proposed approach, we applied our algorithm to a data set where the true temporal order and rotational orientation of the images were known {\em a priori}.
%
This data set was obtained through live imaging of a {\it Drosophila} embryo during the twenty minutes spanning the late stages of cellularization through early gastrulation.
%
During this time window, the ventral furrow is formed, where the ventral side buckles towards the center of the embryo, internalizing the future muscle cells and forming a characteristic ``omega'' shape.
%
Germband extension then causes cells from the ventral side to move towards the posterior pole of the embryo, and then wrap around to the dorsal side \cite{leptin2005gastrulation}.
%
At the end of this process, cells which were originally on the ventral and posterior side of the embryo find themselves on the dorsal side, causing a similar ``omega'' to appear on the dorsal side.

@@

We used vector diffusion maps to register and order the images. 
%
\fig~\ref{subfig:movie_registered_ordered} shows the images from \fig~\ref{subfig:movie_unregistered_unordered}, now registered and ordered; the real time for each frame is also indicated.
%
With a small number of exceptions, the recovered ordering is consistent with the real time dynamics. 
%
\fig~\ref{subfig:movie_angle_corr} and \fig~\ref{subfig:movie_rank_corr}  show the correlations between the recovered and true angles and ranks, respectively, for the entire data set. 
%
Both the angles and the ranks are recovered with a high degree of accuracy.

@@

In general, to accurately recover the temporal order, the vector diffusion maps algorithm requires a sufficient amount of data so that the underlying one-dimensional curve (i.e., the curve in \fig~\ref{subfig:dmaps1}) is well-sampled.
%
At the same time, the accuracy of the recovered alignments is a function of the signal-to-noise ratio in the pairwise alignments, as well as the size of the data set \cite{singer2011angular}.
%
Clearly, $40$ images are sufficient to accurately recover the rotational orientation and temporal order  in this specific developmental time frame.
%
To quantitatively assess how our methods perform as a function of sample size, we used bootstrap sampling to compute the rank correlation as a function of the size of the data set.
%
Subsets of images were randomly selected from the $40$-image data set, and then registered and ordered. 
%
\fig~\ref{subfig:movie_bootstrap} shows the median rank correlation coefficient as a function of the number of images in the subset.
%
As expected, the accuracy in the recovered ordering increases with the number of images. 
%
Because our particular data set contains relatively low noise levels, few images were required to obtain accurate rotational alignments and the errors in the recovered angles were consistently less than $10^{\circ}$.
%
Therefore, the algorithm accurately and robustly registers and orders a model data set where both the rotation angles and real times are independently known.

@@

We have analyzed how our algorithm performs on a model data set, where all images come from a single embryo. 
%
Often in practice, each image comes from a different embryo, and the largest source of noise in the considered data set arises from embryo-to-embryo variability.
%
To demonstrate that our methods are robust to such noise sources,
we constructed a synthetic time course data set by selecting an image from a random live imaging experiment at each time point.
%
The resulting data set is therefore spatially unregistered, scrambled in time, and reflects embryo-to-embryo variability. 
%
The median rank correlation coefficient when ordering such a synthetic time course using our methodology was 0.81, indicating that the algorithm can accurately recover the temporal order even under noisy conditions.

Finally, we applied our approach to a data set where the true rotational orientation and temporal order was not known {\it a priori}.
%
\fig~\ref{subfig:fixed_images_unregistered_unordered} shows selected images from a set of $120$ images (see \fig~\ref{fig:full_fixed_data}) which cover a thirty minute time interval spanning late cellularization through gastrulation.
%
This data set is more complex than the live imaging data sets in that it contains significantly more images, each of which provide information about tissue morphology and the spatial distribution of two regulatory proteins.
%
Each image shows an optical cross-section of a {\em different} vertically oriented embryo at a different rotational orientation and fixed at a different (and unknown) developmental time.
%
The nuclei (gray) were labeled with DAPI, a DNA stain.
%
Embryos were stained with the antibody that recognizes Twist (Twi, shown in green), a transcription factor which specifies the cells of the future muscle tissue.
%
Another signal is provided by the phosphorylated form of the extracellular signal regulated kinase (dpERK, shown in red), an enzyme that, in this context, specifies a subset of neuronal cells.

@@

We can use prior knowledge about the developmental system to evaluate the quality of our registration and ordering. 
%
The Twi signal is known to form a single peak at the ventralmost point of the embryo. 
%
We found that the standard deviation in the location of this peak in the set of registered images was 8$^\circ$,
indicating that the algorithm successfully aligns the ventralmost points of the images. 
%
Because the developmental time of each embryo cannot be easily estimated, we have few options for evaluating the quality of our temporal ordering. 
%
We compared the ordering obtained from vector diffusion maps to the ordering provided by a trained embryologist who is knowledgeable about the developmental progression and the important image features.  
%
The ranks from the ordering provided by the embryologist, which we will refer to as the ``expert rank'', are indicated for the images in \fig~\ref{subfig:fixed_images_registered_ordered} , and the rank correlation (see \fig~\ref{subfig:fixed_images_rank_corr}) shows that our ordering is consistent with the expert ordering.

Manually ordering the images is nontrivial for researchers who are unfamiliar with the developmental progression, and can be tedious and time-consuming for those who are.
%
In contrast, our method requires relatively little computational time; registering and ordering this fixed imaging data set took $15$~seconds on an Intel Core i7 2.93 GHz processor. %for $100$ images of $100 \times 100$ pixels, using 10$^{\circ}$ rotational discretizations.
%
The computational time is a function of the number of images in the data set, the number of pixels in each point, and the angular resolution to compute the pairwise rotations (see \SI). 
%
Empirically, the CPU time required was $\mathcal{O}(n^{1.5})$ in the number of images, and $\mathcal{O}(n)$ in the number of pixels and the angular resolution.
%
Furthermore, the computation of the pairwise rotational alignments, which is the most time-intensive portion of the calculation, is trivially parallelizable.
%
The requisite user intervention and parameter tuning required for our method is relatively minimal.
%
Images must first be preprocessed so that the Euclidean distance between the pixels is informative.
%
Two algorithmic parameters, the angular discretization to compute the pairwise alignments and the diffusion maps kernel scale which determines which data points are ``close'' (see ~\fig~\ref{subfig:dmaps1} and \SI), must also be defined.
%
We found that the results are robust to both of these parameters, and that consistent parameter values served well for all data sets presented here. 
%
%In our examples, we use 10$^{\circ}$ angle discretizations for the pairwise alignments, and one-tenth of the median pairwise distance for the kernel scale. 
% (5$^{\circ}$--30$^{\circ}$ angle discretizations and kernel scales of one-fifth to one-twentieth of the median pairwise distances all yielded acceptable results)
%
The tasks of image preprocessing and parameter selection are relatively simple compared to manual registration and ordering of images, and so this methodology is promising for much larger imaging data sets which are impractical to evaluate manually.

@@

We presented a unified approach to temporal ordering and registration of images in studies of developmental dynamics. 
%
To the best of our knowledge, algorithmic approaches to these two tasks have been explored largely independently of each other. 
%
In particular, temporal ordering of imaging data sets was done with a significant amount of human supervision and using registered images as a starting point \cite{yuan2014automated, surkova2008characterization}.  
%
Temporal ordering of large-scale data which did not require registration was done in the context of molecular profiling studies, in which data are vectors describing the expression levels of different mRNA.
%
Automated temporal ordering of such data sets was accomplished by first projecting the data onto a low-dimensional subspace spanned by the leading principal or independent components, and then solving a traveling salesman problem or constructing a minimum spanning tree on the projected data to order multiple snapshots \cite{anavy2014blind, trapnell2014dynamics, gupta2008extracting}. 
%
The presented approach addresses image registration and ordering in a single step and orders the data without first projecting them on a low-dimensional subspace.
%; vector diffusion maps can also order images when global registration of the entire data set is topologically impossible \cite{zhao2014rotationally}.
%
%The task of image registration has been widely studied \cite{zitova2003image}, for applications such as face recognition \cite{rowley1998rotation}, medical image registration \cite{hajnal2010medical}, and texture classification \cite{greenspan1994rotation}.
%
In contrast to most of the existing registration approaches which rely on the knowledge of appropriate landmarks in the images \cite{ian1998statistical} (such as the eyes in face recognition applications \cite{zhao2003face}), algorithms based on angular synchronization can register images even in the absence of such information, making them relevant for a wide variety of applications. 
%
Although we have presented results for a specific type of data (cross-sectional images taken along the dorsoventral axis of developing {\em Drosophila} embryos), our methodology only relies on computing distances between images and may be readily extended to many other imaging data sets.

Previously, angular synchronization and vector diffusion maps have been used to reconstruct molecular shapes from cryo-electron microscopy images \cite{singer2012vector, zhao2014rotationally, singer2011viewing}.
%
Because of high levels of instrument noise in these data, thousands of images were needed for successful shape reconstruction. 
%
Based on the presented results, we expect that much smaller data sets may be sufficient for successful reconstruction of developmental trajectories from snapshots of fixed embryos.
%
In general, the size of the data set required for accurate registration and ordering is a function of the instrument noise, interembryo variablity, and the complexity of the developmental dynamics.
%
%Here, we have used vector diffusion maps to register the images with respect to rotations; however, one could factor out other sources of variability, such as translations, reflections \cite{singer2012vector, goemans1995improved, bandeira2013cheeger} or dilations.

Already at this point, the rate of collection of imaging data has surpassed the rate of manual image analysis.
%
This necessitates the development of automated methodologies to organize such large data sets;
such methods should be sufficiently general to address a wide variety of applications
%
Vector diffusion maps allow us to automatically register images, an essential task for many applications.
%
Simultaneously, the algorithm provides us with parameters to describe each image.
%
In the examples presented here, we have focused on ordering the images in time using the first vector diffusion maps coordinate.
%
In general, we can recover several parameters which concisely and comprehensively describe data set.
%
This parameterization can then be used for typical data analysis tasks, such as outlier detection and model fitting.
%
Data which comes from several different genetic backgrounds can be clustered according to their spatial expression patterns, and clustering can also be used to separate images into different developmental stages. 
%
Images taken from different viewing directions can be analyzed, as the vector diffusion maps parameterization will organize the images according to the viewing angle.
%
Another direction for future work is related to the joint analysis of data sets provided by different imaging approaches, such as merging live imaging data of tissue morphogenesis with snapshots of cell signaling and gene expression from fixed embryos \cite{krzic2012multiview, ichikawa2014live, rubel2010coupling}.  
%
Given the rapidly increasing volumes of imaging data from studies of multiple developmental systems, we expect that dimensionality reduction approaches discussed in this work will be increasingly useful for biologists and motivate future applications and algorithmic advances.

@@

\section{Software}
%
All algorithms and analysis were implemented in MATLAB\textsuperscript{\textregistered} (R2013b, The MathWorks, Natick, Massachusetts).
%
Code is available at .....

@@

\bibliographystyle{pnas2}
\bibliography{../image_analysis_paper/background_reading/references,../../references/references}

>>>>

\significancetext{
Spatial registration and temporal ordering of images of developing embryos are one of the first steps in the reconstruction of developmental dynamics. 
%
When the number of images is small and the differences between them are easily recognizable, image registration and ordering can be done manually; these tasks become challenging when the data sets are larger and more complex. 
%
Here we show this can be automated and combined in a single
step using recent advances in data mining. 
%
To illustrate this approach, we order and register data from imaging studies of {\it Drosophila}embryogenesis. }

@@

Imaging studies provide unique insights into the dynamics of pattern formation and morphogenesis in developing tissues.
%
Developmental progress is commonly reconstructed from snapshots of chemical or mechanical processes in fixed embryos.
%
As a first step in these reconstructions, snapshots must be spatially registered and ordered in time.
%
Currently, image registration and ordering is often done manually, requiring a significant amount of expertise with a specific developmental system.
%
However, as the sizes of imaging data sets grow, these tasks become increasingly difficult, especially when the images are noisy and the examined developmental changes are subtle.
%
To address these challenges, we present an automated approach to simultaneously register and temporally order imaging data sets.
%
The approach is based on vector diffusion maps, a broadly applicable manifold learning technique that does not require {\it a priori} knowledge of image features or a parametric model of the developmental dynamics.
%
To illustrate our approach, we register and order data from imaging studies of pattern formation and morphogenesis in  {\it Drosophila} embryos.

@@

We present a robust algorithmic approach to registration and temporal ordering.
%
In contrast to a number of previous methodologies \cite{zitova2003image, rowley1998rotation, hajnal2010medical, greenspan1994rotation, zhao2003face}, our methodology does not rely on the {\em a priori} knowledge of spatial landmarks for registration or markers of developmental progression for temporal ordering.
%
Our approach is based on a manifold learning algorithm (vector diffusion maps \cite{singer2012vector}) which simultaneously addresses the problems of registration and temporal ordering. 
%
This algorithm is one of several nonlinear dimensionality reduction techniques that have been developed over the past decade \cite{Belkin2003, coifman2005geometric, coifman2006geometric, tenenbaum2000global, roweis2000nonlinear}, for
applications ranging from analysis of cryo-electron microscopy (cryo-EM) images of individual molecules  \cite{zhao2014rotationally, singer2011viewing} to face recognition \cite{lafon2006data} and classification of CT scans \cite{fernandez2014diffusion}.
%
Here, the vector diffusion maps algorithm is adapted for the analysis of images of developing embryos in studies of developmental dynamics, with the main objective of revealing stereotypic developmental trajectories from fixed images.
%
To illustrate our approach, we show analyze two data sets from a study of {\it Drosophila} embryogenesis, one of the best experimental models for studies of development \cite{jaeger2012drosophila}.
%
Our first data set, from a live imaging study where the correct rotational orientation and temporal order are independently known, will be used to validate our approach.
%
Our second data set consists of images from fixed embryos where the correct orientation and order is unknown; here, we will show how the algorithm can help uncover developmental dynamics which are not readily apparent.

@@

After removing variability due to rotations, the developmental dynamics may be revealed by ordering the data along the one-dimensional manifold that parameterizes most of the remaining variability in the data.
%
Such a manifold can be discovered using diffusion maps \cite{coifman2005geometric}, a nonlinear dimensionality reduction technique that uncovers an intrinsic parameterization of data that lies on a low-dimensional manifold in high-dimensional space.
%
The idea is illustrated in \fig~\ref{subfig:dmaps1}, where the data are two-dimensional points which lie on a one-dimensional nonlinear curve.
%
We use {\it local} information about the data to find a parameterization which respects the underlying manifold geometry, so that points which are close in high-dimensional space (e.g., images which look similar) are close in our parameterization.
%
This idea of locality is denoted by the color of the edges in \fig~\ref{subfig:dmaps1}:
data points which are close are connected by dark edges, and clearly, the dark edges are more ``informative" about the low-dimensional structure of the data.
%
The color in \fig~\ref{subfig:dmaps2} depicts the one-dimensional parameterization or ordering of the data that we can detect visually.
%
In our working examples, each data point will be of much higher dimension (e.g., a pixelated image), and so we cannot extract this low-dimensional structure visually.
%
Instead, we will use diffusion maps to automatically uncover a parameterization of our high-dimensional data (see \SI).
%
Diffusion maps generalizes directly from one-dimensional nonlinear curves to higher-dimensional manifolds.

@@

To validate the proposed approach, we applied our algorithm to a data set where the true temporal order and rotational orientation of the images were known {\em a priori}.
%
This data set was obtained through live imaging of a vertically oriented {\it Drosophila} embryo during the twenty minutes spanning the late stages of cellularization through early gastrulation.
%
During this time window, the ventral furrow is formed, where the ventral side buckles towards the center of the embryo, internalizing the future muscle cells and forming a characteristic ``omega'' shape.
%
Germband extension then causes cells from the ventral side to move towards the posterior pole of the embryo, and then wrap around to the dorsal side \cite{leptin2005gastrulation}.
%
At the end of this process, cells which were originally on the ventral and posterior side of the embryo find themselves on the dorsal side, causing a similar ``omega'' to appear on the dorsal side.

@@

We used vector diffusion maps to register and order the images. 
%
\fig~\ref{subfig:movie_registered_ordered} shows the images from \fig~\ref{subfig:movie_unregistered_unordered}, now registered and ordered; the real time for each frame is also indicated.
%
With a small number of exceptions, the recovered ordering is consistent with the real time dynamics. 
%
\fig~\ref{subfig:movie_angle_corr} and \fig~\ref{subfig:movie_rank_corr}  show the correlations between the recovered and true angles and ranks, respectively, for the entire data set: 
%
Both the angles and the ranks are recovered with a high degree of accuracy.

@@

In general, to accurately recover the temporal order, the vector diffusion maps algorithm requires a sufficient amount of data so that the underlying one-dimensional curve (i.e., the curve in \fig~\ref{subfig:dmaps1}) is well-sampled.
%
At the same time, the accuracy of the recovered alignments is a function of the signal-to-noise ratio in the pairwise alignments, as well as the size of the data set \cite{singer2011angular}.
%
Clearly, $40$ images are sufficient to accurately recover the rotational orientation and temporal order in this specific developmental time frame.
%
To quantitatively assess how our methods perform as a function of sample size, we used bootstrap sampling to compute the rank correlation as a function of the size of the data set.
%
Subsets of images were randomly selected from the $40$-image data set, and then registered and ordered. 
%
\fig~\ref{subfig:movie_bootstrap} shows the median rank correlation coefficient as a function of the number of images in the subset.
%
As expected, the accuracy in the recovered ordering increases with the number of images. 
% 
At the same time, we see that correct order is recovered even with a relatively small number of snapshots, which should make our approach applicable other studies of developmental processes.
%
Because our particular data set contains relatively low noise levels, few images were required to obtain accurate rotational alignments and the errors in the recovered angles were consistently less than $10^{\circ}$.
%
In summary, the algorithm accurately and robustly registers and orders a model data set where both the rotation angles and real times are independently known.

@@

We have analyzed how our algorithm performs on a model data set, where all images come from a single embryo. 
%
In practice, each image comes from a different embryo, and the largest source of noise in the considered data set arises from embryo-to-embryo variability.
%
To demonstrate that our methods are robust to such noise sources,
we constructed a synthetic time course data set by selecting an image from a random live imaging experiment at each time point.
%
The resulting data set is spatially unregistered, scrambled in time, and reflects embryo-to-embryo variability. 
%
The median rank correlation coefficient when ordering such a synthetic time course using our methodology was 0.81, indicating that the algorithm can accurately recover the temporal order even under noisy conditions.

Finally, we applied our approach to a data set where the true rotational orientation and temporal order was not known {\it a priori}.
%
\fig~\ref{subfig:fixed_images_unregistered_unordered} shows selected images from a set of $120$ images (see \fig~\ref{fig:full_fixed_data}) which cover a thirty minute time interval spanning late cellularization through gastrulation.
%
This data set is more complex than the live imaging data sets in that it contains significantly more images, each of which provides information about tissue morphology and the spatial distribution of two regulatory proteins.
%
Each image shows an optical cross-section of a {\em different} vertically oriented embryo at a different rotational orientation and fixed at a different (and unknown) developmental time.
%
The nuclei (gray) were labeled with DAPI, a DNA stain.
%
Embryos were stained with the antibody that recognizes Twist (Twi, shown in green), a transcription factor which specifies the cells of the future muscle tissue.
%
Another signal is provided by the phosphorylated form of the extracellular signal regulated kinase (dpERK, shown in red), an enzyme that, in this context, specifies a subset of neuronal cells \cite{Lim2013kinetics}.

@@

To evaluate the quality of our registration and ordering, we can use prior knowledge about the developmental system. 
%
The Twi signal is known to form a single peak at the ventralmost point of the embryo. 
%
We found that the standard deviation in the location of this peak in the set of registered images was $\sim$8$^\circ$,
indicating that the algorithm successfully aligns the ventralmost points of the images. 
%
Because the developmental time of each embryo cannot be easily estimated, we have few options for evaluating the quality of our temporal ordering. 
%
We compared the ordering obtained from vector diffusion maps to the ordering provided by a trained embryologist who is knowledgeable about the developmental progression and the important image features.  
%
The ranks from the ordering provided by the embryologist, which we will refer to as the ``expert rank'', are indicated for the images in \fig~\ref{subfig:fixed_images_registered_ordered} , and the rank correlation (see \fig~\ref{subfig:fixed_images_rank_corr}) shows that our ordering is consistent with the expert ordering.

Manually ordering the images is nontrivial for researchers who are unfamiliar with the developmental progression, and can be tedious and time-consuming for those who are.
%
In contrast, our method requires relatively little computational time; registering and ordering this fixed imaging data set took $15$~seconds on an Intel Core i7 2.93 GHz processor. %for $100$ images of $100 \times 100$ pixels, using 10$^{\circ}$ rotational discretizations.
%
The computational time is a function of the number of images in the data set, the number of pixels in each point, and the angular resolution to compute the pairwise rotations (see \SI). 
%
Empirically, the CPU time required was $\mathcal{O}(n^{1.5})$ in the number of images, and $\mathcal{O}(n)$ in the number of pixels and the angular resolution.
%
Furthermore, the computation of the pairwise rotational alignments, which is the most time-intensive portion of the calculation, is trivially parallelizable.
%
The requisite user intervention and parameter tuning required for our method is relatively minimal: Images must first be preprocessed so that the Euclidean distance between the pixels is informative.
%
Two algorithmic parameters, the angular discretization to compute the pairwise alignments and the diffusion maps kernel scale which determines which data points are ``close'' (see ~\fig~\ref{subfig:dmaps1} and \SI), must also be defined.
%
We found that the results are robust to both of these parameters, and that consistent parameter values served well for all data sets presented here. 
%
%In our examples, we use 10$^{\circ}$ angle discretizations for the pairwise alignments, and one-tenth of the median pairwise distance for the kernel scale. 
% (5$^{\circ}$--30$^{\circ}$ angle discretizations and kernel scales of one-fifth to one-twentieth of the median pairwise distances all yielded acceptable results)
%
Overall, the tasks of image preprocessing and parameter selection are relatively simple compared to manual registration and ordering of images, and so this methodology is promising for much larger imaging data sets which are impractical to evaluate manually.

@@

Already at this point, the rate of collection of imaging data has surpassed the rate of manual image analysis.
%
This necessitates the development of automated methodologies to organize such large data sets. 
%
To the best of our knowledge, algorithmic approaches to these two tasks have been explored largely independently of each other. 
%
Temporal ordering of large-scale data which did not require registration was done in the context of molecular profiling studies, in which data are vectors describing the expression levels of different mRNA \cite{anavy2014blind, trapnell2014dynamics, gupta2008extracting}.
%
At the same time, temporal ordering of imaging data sets was done with a significant amount of human supervision and using registered images as a starting point \cite{yuan2014automated, surkova2008characterization}.  
%
%The task of image registration has been widely studied \cite{zitova2003image}, for applications such as face recognition \cite{rowley1998rotation}, medical image registration \cite{hajnal2010medical}, and texture classification \cite{greenspan1994rotation}.
%
In contrast to most of the existing registration approaches which rely on the knowledge of appropriate landmarks in the images \cite{ian1998statistical} (such as the eyes in face recognition applications \cite{zhao2003face}), algorithms based on angular synchronization can register images even in the absence of such information, making them relevant for a wide variety of applications. 
%
Although we have presented results for a specific type of data (cross-sectional images taken along the dorsoventral axis of {\em Drosophila} embryos), our methodology only relies on computing distances between images and may be readily extended to many other data sets.

Angular synchronization and vector diffusion maps have been used to reconstruct molecular shapes from cryo-electron microscopy images \cite{singer2012vector, zhao2014rotationally, singer2011viewing}.
%
Because of high levels of instrument noise in these data, thousands of images were needed for successful shape reconstruction. 
%
Based on the presented results, we expect that much smaller data sets may be sufficient for successful reconstruction of developmental trajectories from snapshots of fixed embryos.
%
In general, the size of the data set required for accurate registration and ordering is a function of the instrument noise, interembryo variability, and the complexity of the developmental dynamics.
%

Vector diffusion maps allow us to automatically register images, an essential task for many applications.
%
Simultaneously, the algorithm provides us with parameters to describe each image.
%
In the examples presented here, we have focused on ordering the images in time using the first vector diffusion maps coordinate.
%
In general, we can recover several parameters which concisely and comprehensively describe data set.
%
This parameterization can then be used for typical data analysis tasks, such as outlier detection and model fitting.
%
Data which comes from several different genetic backgrounds can be clustered according to their spatial expression patterns, and clustering can also be used to separate images into different developmental stages. 
%
Furthermore, images taken from different viewing directions can be analyzed, as the vector diffusion maps parameterization will organize the images according to the viewing angle.
%
Another direction for future work is related to the joint analysis of data sets provided by different imaging approaches, such as merging live imaging data of tissue morphogenesis with snapshots of cell signaling and gene expression from fixed embryos \cite{krzic2012multiview, ichikawa2014live, rubel2010coupling}.  
%
Given the rapidly increasing volumes of imaging data from studies of multiple developmental systems, we expect that dimensionality reduction approaches discussed in this work will be increasingly useful for biologists and motivate future applications and algorithmic advances.

@@

\section{Software}
%
All algorithms and analysis were implemented in MATLAB\textsuperscript{\textregistered} (R2013b, The MathWorks, Natick, Massachusetts).
%
Code is available at \texttt{genomics.princeton.edu/stas/publications.html} under ``Codes and Data''.

@@

\bibliographystyle{pnas2}
\bibliography{../../references/references,../image_analysis_paper/background_reading/references}
